Changes in Patterns of Geometric Sign Groupings as a Model of Social Dynamics Across Aurignacian Phases

### 2025-04-14 endofquarterpaper.qmd

Similar analyses of personal ornament types from the Aurignacian technocomplex identified fourteen geographically cohesive groups [@Vanhaeren_d]. Drawing on ethnographic studies that show how body decoration and ornamentation indicates ethno-linguistic identity, @Vanhaeren_d used Aurignacian personal ornamentation as a proxy for ethno-linguistic diversity. They analyzed 157 distinct ornament types from 98 Aurignacian sites in Europe and the Near East. Geographically cohesive groups sharing similar ornament type associations were identified by seriation, correspondence, and GIS analyses. These groups sweep counter-clockwise throughout western France, northern Spain, the Pyrenees, and the Mediterranean region. @Vanhaeren_d interpret the variation in personal ornament types as evidence of long-lasting cultural differences and as robustly establishing the ethno-linguistic diversity of the Aurignacian.

@@

The first aim of our study was to determine how the grouping dynamics of geometric signs changed across four phases of the Aurignacian, specifically in terms of diversity, size, connectivity, and boundaries. The results from @fig-seriation show that each phase can be divided into two main cultural groups, with sign type diversity serving as a major dividing factor, and that group size varied from phase to phase, with a large increase in between the Transitional phase and the Proto-Aurignacian. @fig-neighbor-net shows that group inter-connectivity and intra-connectivity varies from phase to phase, with weak intra-group connections and no inter-group connections in the Transitional phase, to stronger intraconnections and interconnections starting to form in the Proto-Aurignacian, to extremely strong connections (both between and within groups) in the Early Aurignacian, to weaker connections in the Evolved Aurignacian. The modularity tests, as shown in @tbl-strength-table, show that the groupings are strongest in the Transitional period and weakest in the Early Aurignacian, and the perMANOVA tests tell us that the Transitional period experiences the highest level of difference between groups, while the Early Aurignacian experiences the least. The Mantel tests, as shown in @tbl-mantel-table, tell us that the sign distribution in the Proto-Aurignacian experiences the lowest correlation with geographic distance, while the Early Aurignacian experiences the highest correlation.

@@

The geographic boundaries of the groups change from phase to phase (@fig-group-map). In the Transitional phase, sites are widely spread out across Western Europe, and tend to be located closely to the other sites in their groups. The exception is Hohle Fels, which, instead of being located further west with the rest of the sites in the broader group, is located further east, closer to the sites in the restricted group, and almost overlapping with Geissenklösterle. In the Proto-Aurignacian, we see a slightly narrower distribution of sites westward but a spread northward, and more overlapping of sites between groups. In the Early Aurignacian, we see a reduced distribution, with sites clustered closer together. In the Evolved Aurignacian, we see a wider distribution and more distance between sites.

>>>>

Similar analyses of personal ornament types from the Aurignacian technocomplex identified fourteen geographically cohesive groups [@Vanhaeren_d]. Drawing on ethnographic studies that show how body decoration and ornamentation indicates ethno-linguistic identity, @Vanhaeren_d used Aurignacian personal ornamentation as a proxy for ethno-linguistic diversity. The variation in personal ornament types is interpreted as evidence of long-lasting cultural differences and as robustly establishing the ethno-linguistic diversity of the Aurignacian. 157 distinct ornament types from 98 Aurignacian sites in Europe and the Near East were analyzed, identifying geographically cohesive groups sharing similar ornament type associations by seriation, correspondence, and GIS analyses. These groups sweep counter-clockwise throughout western France, northern Spain, the Pyrenees, and the Mediterranean region.

@@

Our first aim was to determine how the grouping dynamics of geometric signs changed across four phases of the Aurignacian, specifically in terms of diversity, size, connectivity, and boundaries. The results from @fig-seriation show that each phase can be divided into two main cultural groups, with sign type diversity serving as a major dividing factor, and that group size varied from phase to phase, with a large increase in between the Transitional phase and the Proto-Aurignacian. @fig-neighbor-net shows that group inter-connectivity and intra-connectivity varies from phase to phase, with weak intra-group connections and no inter-group connections in the Transitional phase, to stronger intraconnections and interconnections starting to form in the Proto-Aurignacian, to extremely strong connections (both between and within groups) in the Early Aurignacian, to weaker connections in the Evolved Aurignacian. The modularity tests, as shown in @tbl-strength-table, show that the groupings are most distinct in the Transitional period and least distinct in the Early Aurignacian. The perMANOVA tests show that the Transitional period experiences the highest level of difference between groups, while the Early Aurignacian has the least. The Mantel tests, as shown in @tbl-mantel-table, tell us that the sign distribution in the Proto-Aurignacian experiences the lowest correlation with geographic distance, while the Early Aurignacian has the highest correlation.

@@

The geographic boundaries of the groups change from phase to phase (@fig-group-map). In the Transitional phase, sites are widely spread out across Western Europe. This wider distribution may be a factor in weak connections seen in @fig-neighbor-net. In the Proto-Aurignacian, we see a slightly narrower distribution of sites westward but a spread northward, and more overlapping of sites between groups, which may be associated with the stronger connectivity seen there. In the Early Aurignacian, we see a reduced distribution, with sites clustered closer together, which may be associated with the increased connectivity in that phase. In the Evolved Aurignacian, we see a wider distribution and more distance between sites, which may be associated with the reduced connectivity seen in that phase.
